Racial and Ethnic Differences in ETI

### 2023-09-25 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti.qmd
initial survival analysis done

```{r setup}
#| include: false
# Libraries
library(Hmisc)
library(tidyverse)
library(gtsummary)
library(lmerTest)
library(nlme)
library(broom.mixed)
library(emmeans)
library(gt)
library(performance)
library(mcp)
```

```{r data cleaning}
#| include: false
# source("/Users/timvigers/GitHub/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/create_race_eti_analysis_dataset.R")
load("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/analysis_dataset.RData")
```

@@

For the following analysis, the start time ($t=0$) was the approximate date that each person became eligible for ETI. Because we don't have exact dates of birth, I calculated an approximate DOB based on age and encounter dates.

```{r}
# Create survival dataset
surv_dat <- encounter %>%
  mutate(stop = as.numeric(encounterdate - eti_elig_date)) %>%
  filter(stop >= 0) %>%
  group_by(eDWID) %>%
  mutate(start = lag(stop)) %>%
  select(eDWID, race, encounterdate, encounterage, eti_elig_date, Vx445comb, start, stop) %>%
  filter(stop <= first(stop[Vx445comb == "Yes"]))
surv_dat$Vx445comb <- as.numeric(factor(surv_dat$Vx445comb, levels = c("No", "Yes")))
# Filter for KM
km_dat <- surv_dat %>%
  group_by(eDWID) %>%
  slice_max(stop)
kmfit <- survfit(Surv(stop, Vx445comb) ~ race, data = km_dat[km_dat$race != "Asian", ])
autoplot(kmfit) + theme_bw()
# Cox PH model
fit <- coxph(Surv(start, stop, Vx445comb) ~ race, id = eDWID, data = surv_dat)
```

@@

```{r}
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_line(mapping = aes(x = ReviewYear, y = mean_ppFEV, group = eDWID), alpha = 0.005, linewidth = 0.5) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean, color = race)) +
  geom_line(mapping = aes(x = ReviewYear, y = emmean, group = race, color = race)) +
  geom_errorbar(mapping = aes(x = ReviewYear, ymin = emmean - SE, ymax = emmean + SE, color = race), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity")
```

```{r}
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_line(mapping = aes(x = ReviewYear, y = mean_ppFEV, group = eDWID), alpha = 0.005, linewidth = 0.5) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean, color = race)) +
  geom_line(mapping = aes(x = ReviewYear, y = emmean, group = race, color = race)) +
  geom_errorbar(mapping = aes(x = ReviewYear, ymin = emmean - SE, ymax = emmean + SE, color = race), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity") +
  facet_wrap(~race)
```

>>>>

```{r setup}
#| include: false
# Libraries
library(Hmisc)
library(tidyverse)
library(gtsummary)
library(lmerTest)
library(nlme)
library(broom.mixed)
library(emmeans)
library(gt)
library(performance)
library(mcp)
library(ggsurvfit)
```

```{r data cleaning}
#| include: false
# source("/Users/timvigers/GitHub/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/create_race_eti_analysis_dataset.R")
# load("UCD/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/analysis_dataset.RData")
load("/Users/timvigers/Dropbox/Work/Vigers/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/analysis_dataset.RData")
```

@@

For the following analysis, the start time ($t=0$) was the approximate date that each person became eligible for ETI. Because we don't have exact dates of birth, I calculated an approximate DOB based on age and encounter dates. Encounters prior to ETI eligibility were excluded, and only time to first encounter on ETI was considered (i.e. we did not fit a multi-state model to account for those who started and then stopped the medication).

```{r}
# Create survival dataset
surv_dat <- encounter %>%
  mutate(stop = as.numeric(encounterdate - eti_elig_date)) %>%
  filter(stop >= 0) %>%
  group_by(eDWID) %>%
  mutate(start = lag(stop, default = 0)) %>%
  select(eDWID, race, encounterdate, encounterage, eti_elig_date, Vx445comb, start, stop) %>%
  filter(stop <= first(stop[Vx445comb == "Yes"]))
surv_dat$Vx445comb <- as.numeric(factor(surv_dat$Vx445comb, levels = c("No", "Yes"))) - 1
# Fit
fit <- survfit(Surv(start, stop, Vx445comb, type = "interval") ~ race,
  id = eDWID, data = surv_dat
)
# Plot
names(fit$strata) <- gsub("race=", "", names(fit$strata))
fit %>%
  ggsurvfit(type = "risk") +
  add_confidence_interval() +
  add_risktable(
    risktable_stats =
      c("{round(n.risk)}", "{round(cum.event)}"),
    stats_label = c("Eligible", "Num. Started")
  ) +
  add_risktable_strata_symbol() +
  add_quantile() +
  ylab("Proportion on Vx445comb") + xlab("Days From Eligibility")
```

Without Asian subgroup for clearer plot:

```{r}
fit <- survfit(Surv(start, stop, Vx445comb, type = "interval") ~ race,
  id = eDWID, data = surv_dat[surv_dat$race != "Asian", ]
)
# Plot
names(fit$strata) <- gsub("race=", "", names(fit$strata))
fit %>%
  ggsurvfit(type = "risk") +
  add_confidence_interval() +
  add_risktable(
    risktable_stats =
      c("{round(n.risk)}", "{round(cum.event)}"),
    stats_label = c("Eligible", "Num. Started")
  ) +
  add_risktable_strata_symbol() +
  add_quantile() +
  ylab("Proportion on Vx445comb") + xlab("Days From Eligibility")
```

@@

```{r}
#| warning: false
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_line(mapping = aes(x = ReviewYear, y = mean_ppFEV, group = eDWID), alpha = 0.005, linewidth = 0.5) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean, color = race)) +
  geom_line(mapping = aes(x = ReviewYear, y = emmean, group = race, color = race)) +
  geom_errorbar(mapping = aes(x = ReviewYear, ymin = emmean - SE, ymax = emmean + SE, color = race), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity")
```

```{r}
#| warning: false
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_line(mapping = aes(x = ReviewYear, y = mean_ppFEV, group = eDWID), alpha = 0.005, linewidth = 0.5) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean, color = race)) +
  geom_line(mapping = aes(x = ReviewYear, y = emmean, group = race, color = race)) +
  geom_errorbar(mapping = aes(x = ReviewYear, ymin = emmean - SE, ymax = emmean + SE, color = race), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity") +
  facet_wrap(~race)
```
